---
title: "Areal Data Overview"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE)
library(tidyverse)
library(mnormt)
library(spdep)
set.seed(03202021)
```


## Proximity Matrix
Similar to the distance matrix with point-reference data, a proximity matrix $W$ is used to model areal data.

\vfill

Given measurements $Y_i, \dots, Y_n$ associated with areal units $1, \dots, n$, the elements of $W$, $w_{ij}$ connect units $i$ and $j$

\vfill

Common values for $w_{ij}$ are
$$w_{ij} =\begin{cases} 1 & \text{if i and j are adjacent} \\ 
0& \mbox{otherwise (or if i=j) } 
\end{cases}$$

\vfill

#### Grid Example
Create an adjacency matrix with diagonal neigbors

\vfill

Create an adjacency matrix without diagonal neigbors

```{r, echo = F}
d=data.frame(xmin=c(0.5,0.5,0.5,-.5,-.5,-.5,-1.5,-1.5,-1.5),
             xmax=c(1.5,1.5,1.5,.5,.5,.5,-.5,-.5,-.5),
             ymin=rep(c(.5,-.5,-1.5), 3), 
             ymax=rep(c(1.5,.5,-.5), 3),
             id=c(1,2,3,4,5,6,7,8,9))
ggplot() + 
  scale_x_continuous(name="x") + 
  scale_y_continuous(name="y") +
  geom_rect(data=d, mapping=aes(xmin=xmin, xmax=xmax, ymin=ymin, ymax=ymax), color="black", alpha=0.05) +
  geom_text(data=d, aes(x=xmin+(xmax-xmin)/2, y=ymin+(ymax-ymin)/2, label=id), size=4) +
  theme_minimal()

```

\newpage

## Spatial Association

There are two common statistics used for assessing spatial association:
Moran's I and Geary's C.

\vfill

Moran's I
$$I =\frac{n \sum_i \sum_j w_{ij} (Y_i - \bar{Y})(Y_j -\bar{Y})}{(\sum_{i\neq j \;w_{ij}})\sum_i(Y_i - \bar{Y})^2}$$

\vfill

Moran's I is analogous to correlation, where values close to 1 exhibit spatial clustering and values near -1 show spatial regularity (checkerboard effect).

\vfill

Geary's C
$$C=\frac{(n-1)\sum_i \sum_j w_{ij}(Y_i-Y_j)^2}{2(\sum_{i \neq j \; w_{ij}})\sum_i (Y_i - \bar{Y})^2}$$

\vfill

Geary's C is more similar to a variogram (has a connection to Durbin-Watson in 1-D). The statistics ranges from 0 to 2; values close to 2 exhibit regularity and values close to 1 show clustering.

\vfill
\newpage

## Spatial Association Exercise

Consider the following scenarios and use the following 4-by-4 grid

```{r, echo = F}
d4 <- tibble(xmin = rep(c(3.5, 2.5, 1.5, 0.5), each = 4),
             x = rep(4:1, each =4),      
             xmax = rep(c(3.5, 2.5, 1.5, 0.5), each = 4) +1,
             ymin = rep(c(3.5, 2.5, 1.5, 0.5), 4), 
             y = rep(4:1, 4),
             ymax = rep(c(3.5, 2.5, 1.5, 0.5), 4) +1,
             rpos = rep(4:1, 4),
             cpos = rep(1:4, each = 4),
             id=16:1)
ggplot() + 
  scale_x_continuous(name="column") + 
  scale_y_continuous(name="row",breaks = 1:4,labels = c('4','3','2', '1') ) +
  geom_rect(data=d4, mapping=aes(xmin=xmin, xmax=xmax, ymin=ymin, ymax=ymax), color="black", alpha=0.05) +
  geom_text(data=d4, aes(x=xmin+(xmax-xmin)/2, y=ymin+(ymax-ymin)/2, label=id), size=4) +
  theme_minimal()
```

and proximity matrix 

```{r}
W <- matrix(0, 16, 16)
for (i in 1:16){
  W[i,] <- as.numeric((d4$rpos[i] == d4$rpos & (abs(d4$cpos[i] - d4$cpos) == 1)) | 
                        (d4$cpos[i] == d4$cpos & (abs(d4$rpos[i] - d4$rpos) == 1)))
}
head(W)
```
\vfill

for each scenario plot the grid, calculate I and G, along with permutation-based p-values.

\newpage

1. Simulate data where the responses are i.i.d. N(0,1). 

```{r, echo = T}
d4$z <- rnorm(16)

ggplot() + 
  geom_tile(data=d4, mapping=aes(x = x, y = y, fill = z)) +
  geom_text(data=d4, aes(x=xmin+(xmax-xmin)/2, y=ymin+(ymax-ymin)/2, label=id), size=4) +
  theme_minimal() + scale_fill_gradient2(midpoint=0, low="blue", mid="white",
                     high="red", space ="Lab" )

moran.test(d4$z, mat2listw(W), alternative = 'two.sided')
#moran.plot(d4$z, mat2listw(W))
geary.test(d4$z, mat2listw(W), alternative = 'two.sided')
```


\vfill

\newpage

2. Simulate data and calculate I and G for a 4-by-4 grid with a chess board approach, where "black squares" $\sim N(-2,1)$ and "white squares" $\sim N(2,1)$.

```{r, echo = T}

d4$z2 <- 0

for (i in 1:16){
  if ((d4$rpos[i] + d4$cpos[i])%% 2 == 1) {
    d4$z2[i] <- rnorm(1, mean = 3)
  } else {
    d4$z2[i] <- rnorm(1, mean = -3)
  }
}

ggplot() + 
  geom_tile(data=d4, mapping=aes(x = x, y = y, fill = z2)) +
  geom_text(data=d4, aes(x=xmin+(xmax-xmin)/2, y=ymin+(ymax-ymin)/2, label=id), size=4) +
  theme_minimal() + scale_fill_gradient2(midpoint=0, low="blue", mid="white",
                     high="red", space ="Lab" )


moran.test(d4$z2, mat2listw(W), alternative = 'two.sided')
geary.test(d4$z2, mat2listw(W), alternative = 'two.sided')
```

\vfill

\newpage

3. Simulate multivariate normal response on a 4-by-4 grid where $y \sim N(0, (I- \rho W)^{-1})$, where $\rho = .3$ is a correlation parameter and $W$ is a proximity matrix.


```{r}
d4$z3 <- mnormt::rmnorm(1, mean = 0, varcov = solve(diag(16) -  .3 * W))

ggplot() + 
  geom_tile(data=d4, mapping=aes(x = x, y = y, fill = z3)) +
  geom_text(data=d4, aes(x=xmin+(xmax-xmin)/2, y=ymin+(ymax-ymin)/2, label=id), size=4) +
  theme_minimal() + scale_fill_gradient2(midpoint=0, low="blue", mid="white",
                     high="red", space ="Lab" )


moran.test(d4$z3, mat2listw(W), alternative = 'two.sided')
geary.test(d4$z3, mat2listw(W), alternative = 'two.sided')
```


\vfill
